Reasoning About a Machine with Local Capabilities

### 2019-02-10 toplas/paper.tex
[TOPLAS] Update capability safety section

\subsection{Safety of the Capability Machine}
With the logical relation defined, we can now state the fundamental theorem of
our logical relation: a strong theorem that formalizes the guarantees offered by
the capability machine. Essentially, it says a capability that only grants safe
authority is capability safe as a program counter.
\begin{theorem}[Fundamental Theorem]
  \label{thm:ftlr}
  If one of the following holds:
  \begin{align*}
      & \bullet
        \begin{aligned}[t]
        &\perm = \exec \text{ and }\npair{(\start,\addrend)} \in \readCond{}(\gl)(W)
      \end{aligned} \\
    & \bullet 
      \begin{aligned}[t]
        &\perm = \rwx \text{ and } \npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{ and }\\
        &\npair{(\start,\addrend)} \in \writeCond{}(\iota^\nwl,\gl)(W)
      \end{aligned} \\
    & \bullet 
      \begin{aligned}[t]
        &\perm = \rwlx \text{ and }\npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{ and }\\
        &\npair{(\start,\addrend)} \in \writeCond{}(\iota^\pwl,\gl)(W),
      \end{aligned}
  \end{align*}
  then $\npair{((\perm,\gl),\start,\addrend,\addr)} \in \stder(W)$
\end{theorem}
\begin{toplas}
  \begin{proof}[Proof sketch]
    Induction over $n$. By definition of $\stder(W)$, show
    \[
      \npair[n']{(\reg\update{\pcreg}{((\perm,\gl),\start,\addrend,\addr)},\ms)}
      \in \observations(W)
    \]
    assuming $n' \leq n$, $\npair[n']{\reg} \in \stdrr(W)$, and
    $\heapSat[\ms]{n'}{W}$. By definition of $\observations$ let $\ms_f$,
    $\mem'$, and $i \leq n'$ be given and for
    $\Phi = (\reg\update{\pcreg}{((\perm,\gl),\start,\addrend,\addr)},\ms \uplus
    \ms_f)$ assume $\Phi \step[i]
    (\halted,\mem')$ and show there exists $W' \futurestr W$ that part of
    $\mem'$ satisfies. First observe that $i \neq 0$ as $\Phi$ is a non-halted
    configuration, so $\Phi$ takes at least one step, i.e.\ $\Phi \step
    \Phi' \step[i-1] (\halted,\mem')$.

    The rest of the proof considers the different ways the step $\Phi \step
    \Phi'$ could have occurred, depending on the instruction being executed. For
    each of these cases, we argue (1) that $\Phi'$ is consistent with the world
    in the sense that the register-file and memory still respect the world and
    (2) that the rest of the execution respects the world. Depending on where
    the $\pcreg$ in $\Phi'$ comes from, the second result is proven in one of
    two ways. If the step to $\Phi'$ was a jump, then the new $\pcreg$ is one of
    the safe values in $\Phi$'s registers and the value relation can be used to
    argue that those can be safely jumped to. On the other hand, if the $\pcreg$
    was just incremented, then we can apply the induction hypothesis.

    In order to argue (1), we consider what configurations the operational
    semantics allows us to get to from the initial state. If we consider the
    memory in $\Phi'$, then it either (a) remains unchanged or (b) one address
    has been updated and in that case, the register-file contains an appropriate
    capability for writing. The latter occurs when the instruction being
    executed is a $\texttt{store}$, and otherwise the former. In case (a), we
    can conclude that the memory still respects the world just by downwards
    closure of memory satisfaction. In case (b), we use an auxiliary lemma that
    uses the safety of the write capability used by the $\texttt{store}$
    instruction to show that the updated memory satisfies the world. To show
    that the updated register-file is safe, we consider the changes made to it
    by all instructions in separate lemmas and show that they all preserve
    safety of the register file.

@@

The Fundamental Theorem can be understood as a general expression of the
guarantees offered by the capability machine, an instance of a general property
called capability safety~\citep{Devriese:2016ObjCap,Maffeis2010OC}. To
understand this, consider that the theorem says the capability
$((\perm,\gl),\start,\addrend,\addr)$ is safe as a program counter, without any
assumption about what instructions it actually points to (the only assumptions
we have are about the read or write authority that it carries). As such, the
theorem expresses the capability safety of the machine, which guarantees that
\emph{any} instruction is fine and will not be able to go beyond the authority
of the values it has access to. We demonstrate this in
\sectionname~\ref{sec:examples} where Theorem~\ref{thm:ftlr} is used to reason about
capabilities that point to arbitrary instructions. The relation between
Theorem~\ref{thm:ftlr} and local-state encapsulation and control-flow
correctness, will also be shown by example in \sectionname~\ref{sec:examples} as the
examples depend on these properties for correctness.
% See the technical appendix~\citep{technical_appendix} for a detailed
% proof (by induction over the step-index $n$) of the theorem.

>>>>

\subsection{Capability Machine Safety}
With the logical relation defined, we can now state the fundamental theorem of
our logical relation: a strong theorem that formalizes the guarantees offered by
the capability machine. Essentially, it says a capability that only grants safe
authority is capability safe as a program counter.
\begin{theorem}[Fundamental Theorem]
  \label{thm:ftlr}
  If one of the following holds:
  \begin{align*}
      & \bullet
        \begin{aligned}[t]
        &\perm = \exec \text{ and }\npair{(\start,\addrend)} \in \readCond{}(\gl)(W)
      \end{aligned} \\
    & \bullet 
      \begin{aligned}[t]
        &\perm = \rwx \text{ and } \npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{ and }\\
        &\npair{(\start,\addrend)} \in \writeCond{}(\iota^\nwl,\gl)(W)
      \end{aligned} \\
    & \bullet 
      \begin{aligned}[t]
        &\perm = \rwlx \text{ and }\npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{ and }\\
        &\npair{(\start,\addrend)} \in \writeCond{}(\iota^\pwl,\gl)(W),
      \end{aligned}
  \end{align*}
  then $\npair{((\perm,\gl),\start,\addrend,\addr)} \in \stder(W)$
\end{theorem}
\begin{toplas}
  \begin{proof}[Proof sketch]
    Induction over $n$. By definition of $\stder(W)$, show
    \[
      \npair[n']{(\reg\update{\pcreg}{((\perm,\gl),\start,\addrend,\addr)},\ms)}
      \in \observations(W)
    \]
    assuming $n' \leq n$, $\npair[n']{\reg} \in \stdrr(W)$, and
    $\heapSat[\ms]{n'}{W}$. By definition of $\observations$ let $\ms_f$,
    $\mem'$, and $i \leq n'$ be given and for
    $\Phi = (\reg\update{\pcreg}{((\perm,\gl),\start,\addrend,\addr)},\ms \uplus
    \ms_f)$ assume $\Phi \step[i]
    (\halted,\mem')$ and show there exists $W' \futurestr W$ that part of
    $\mem'$ satisfies. First observe that $i \neq 0$ as $\Phi$ is a non-halted
    configuration, so $\Phi$ takes at least one step, i.e.\ $\Phi \step
    \Phi' \step[i-1] (\halted,\mem')$.

    The rest of the proof considers the different possibilities for the step $\Phi \step
    \Phi'$, i.e. it considers each of the instructions that could have been executed. For
    each of these cases, we argue that (1) $\Phi'$ is consistent with the world
    in the sense that the register-file and memory still respect the world and
    that (2) the rest of the execution respects the world. Depending on where
    the $\pcreg$ in $\Phi'$ comes from, the second result is proven in one of
    two ways. If the step to $\Phi'$ was a jump, then the new $\pcreg$ is one of
    the safe values in $\Phi$'s registers, and the value relation is used to
    argue that the jump is safe. On the other hand, if the $\pcreg$
    was just incremented, then we can apply the induction hypothesis.

    In order to argue (1), we look at the possible states for $\Phi$ according to the operational semantics.
    If we consider the memory in $\Phi'$, then it either (a) remains unchanged or (b) one address has been updated and the register-file contains an appropriate capability for writing.
    The latter occurs when the executed instruction is $\texttt{store}$.
    Otherwise we are in the former case.
    In case (a), the memory hasn't changed, so we conclude that the memory still respects the world simply by downwards closure of memory satisfaction.
    In case (b), we use an auxiliary lemma that uses the safety of the write capability used by the $\texttt{store}$ instruction to show that the updated memory satisfies the world.
    To show that the updated register-file is safe, we consider the changes made to it by all instructions in separate lemmas and show that they all preserve safety of the register file.

@@

The Fundamental Theorem can be understood as a general expression of the guarantees offered by the capability machine which is an instance of a general property called capability safety~\citep{Devriese:2016ObjCap,Maffeis2010OC}.
The theorem says that an arbitrary capability $((\perm,\gl),\start,\addrend,\addr)$ is safe as a program counter without making any assumptions about what program it points to (the only assumptions we have are about the read or write authority that it carries).
As such, the theorem expresses the capability safety of the machine which guarantees that \emph{any} instruction is safe to execute and will not be able to go beyond the authority of the values it has access to.
We demonstrate this in \sectionname~\ref{sec:examples} where Theorem~\ref{thm:ftlr} is used to reason about capabilities that point to arbitrary instructions.
The relation between Theorem~\ref{thm:ftlr} and local-state encapsulation and control-flow correctness will also be shown by example in \sectionname~\ref{sec:examples} as the examples depend on these properties for correctness.

% See the technical appendix~\citep{technical_appendix} for a detailed
% proof (by induction over the step-index $n$) of the theorem.
